Toward single particle reconstruction without particle picking

### 2018-05-15 TEX/draft3.tex
updates of section 2

The significance of the model bias in cryo--EM was stressed in~\cite{shatsky2009method,henderson2013avoiding} and demonstrated by the ``Einstein from noise'' example.
In this experiment, the image of Einstein is correlated with multiple images of pure noise. The noisy images are aligned to Einstein's image using cross-correlation and then averaged. 
In Figure~\ref{fig:EinsteinFromNoise} we conducted this experiment with different number of noise images. With merely 1000 noise images, Einstein's face is clearly observed in the averaged image. With $10,000$  image, the output is remarkably similar to the template image of Einstein, rather than to pure noise. Crucially, the more noisy images we use, the more severe the model bias towards the template. 
In the context of cryo--EM, this example demonstrates how our prior assumptions about the particle may influence the reconstructed structure.

@@

	\caption{\label{fig:EinsteinFromNoise} Einstein from noise experiment. As a template, we use Einstein's image of size $50\times 50$ pixels. We drawn $n$ images of the same size whose entries are composed of i.i.d.\ Gaussian entries with the same mean and variance as Einstein's image. Each noise image was aligned with respect to the template by the maximal value of their cross correlation. After the alignment, all the images were averaged. As can be seen, the more pure noise images we use, the closer the averaged image to the template.   }	
\end{figure}

@@

Figure~\ref{fig:micro_example} shows a simple example of our problem with  one signal ($K=1$). 
Each panel demonstrates an image of size $250\times 250$, containing 4 occurrences of an Einstein image of size $50\times 50$, appearing in Figure~\ref{fig:Einstein}. 
The left panel shows the data without noise. In the middle panel,  i.i.d.\ Gaussian noise with standard deviation $\sigma=0.5$  was added, however,  Einstein is still easily identified. In this scenario it is likely that standard detection algorithms can locate the copies of the signal within the micrograph. In the right panel, we show the same data swamped in  
i.i.d.\ Gaussian noise with standard deviation  $\sigma=3$.
Clearly, identifying the locations of Einstein in this micrograph is much more challenging. In fact, it can be shown that in the low signal--to--noise (\SNR) regime, 
detection  of individual image occurrences is impossible---even if the true image is known---and therefore particle picking is impossible.  This phenomenon is explained in detail in Appendix~\ref{sec:theory}.

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of data sequences (micrographs) of size $250\times 250$ with additive i.i.d.\ Gaussian noise with variance $\sigma^2$. Each micrograph contains four occurrences of a $50 \times 50$ image of Einstein that appears in Figure~\ref{fig:Einstein}.}	
\end{figure}

@@

In this section we describe  some numerical experiments and show results. The experimental details and the algorithms used are discussed at length in Section~\ref{sec:methods}.

In the first experiment, we estimated Einstein's image of size $50\times 50$ from multiple micrographs of size $4096\times 4096$ pixels, contaminated  with additive i.i.d.\ Gaussian noise with standard deviation $\sigma=3$. An illustrative example of a micrograph appears in the right panel of Figure~\ref{fig:micro_example}.
In each micrograph, multiple occurrences of Einstein's image    were placed, while keeping a separation of  $50$ pixels between the images in each direction. From the micrographs, we estimated the power spectrum, which is equivalent to the second-order autocorrelation. Then, in order to estimate the signal itself, we applied a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR).

As an initial guess, we picked an image of the physicist Issac Newton. If the algorithm was prone to model bias, we would expect to get as an output an image that resembles Newton, similarly to the ``Einstein from noise'' effect. However, the experiment exhibits the desired phenomenon: the more data we collect, the better the reconstruction quality.

Figure~\ref{fig:Einst_example} demonstrates several recovery results for different number of micrographs collected. Figure~\ref{fig:error_per_micro} presents the normalized recovery error and the power spectrum estimation error as a function of the number of micrographs.
To measure the error, we use the root mean square error (RMSE) defined as 
\begin{equation}
\text{RMSE}  := \frac{\|x - \hat{x}\|_{\text{F}}}{\|x \|_{\text{F}}},
\end{equation} 
where $\hat{x}$ is the estimated image and $\|\cdot\|_{\text{F}}$ stands for the Frobenius norm. 
As expected, the RMSE of estimating the power spectrum   decreases linearly with slope $-1/2$ in logarithmic scale. 
\TODO{Few comments: 1. We have a movie in the supplementary material. 2. Does the error also have the right slope? 3. We can put more images to exemplify the progress 4. Need to improve the micrograph's generation code  }

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{Newton}
		\caption{Newton  (model)}
	\end{subfigure}%
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{reconstruction2}
		\caption{\small $P =1024$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{reconstruction20}
		\caption{$P = 1024\times 10$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{reconstruction200}
		\caption{$P = 1024\times 100$}
	\end{subfigure}%
	\caption{\label{fig:Einst_example} Recovery of Einstein's image. The RRR algorithm is initialized with the image of the physicist Issac Newton. We show estimations of Einstein using $P=1024,1024\times 10$ and $1024\times 100$ number of micrographs, each contains $800$ image occurrences on average (to verify). The RMSE is $0.5514$, $0.3494$ and $0.2112$, respectively. The power spectrum estimation RMSE is, respectively, $0.0269 $, $0.0083$ and $0.0027$. \TODO{Can add more images to exemplify the progress}}	
\end{figure}

@@

Results from a similar experiment  with $\sigma=3$ for $K=3$ 1-D signals is presented in Figure~\ref{fig:1Dheterosignals}.
 We define the ratio of the space occupied by the $i$th signal as 
\begin{equation}
\gamma_i = \frac{M_i L}{N},
\end{equation}
where $M_i$ is the number of signal's  occurrences, $L$ is the length of the signal and $N$ is the micrograph length.
These ratios present the \emph{density} of the  $i$th signal in the data.
In the experiment, we do not assume to know these ratios, neither the noise level $\sigma$.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=.5]{Einstein_progress}
\caption{\label{fig:error_per_micro} The left panel shows the recovery error (in log-log scale) of Einstein's image defined in~\eqref{eq:error} as a function of the number of the collected micrographs. The right panel presents the error of estimating the power spectrum of the signal from the data. As expected, in log-log scale, the curve has a $-1/2$ slope.}	
\end{figure}

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{XP_1D_heterogeneous/hetero_example}
	\caption{Red signals are the ground truth (targets) and the blue signals are our estimations. Individual RMSE of the estimates: $0.0239393 / 0.208925 / 0.0335956$. The estimated $\gamma$'s are $0.02574 / 0.01693 / 0.00818$ and the true ones $0.02561 / 0.01707 / 0.00853$. The $\SNR$ is $1/12$. \TODO{to replace with a progress plot`}}
	\label{fig:1Dheterosignals}
\end{figure}

@@

Our results rely on two core assumptions that are not necessarily met by in an SPR experiment.  
First, we modeled the background information as i.i.d.\ additive noise. In practice, the background information may be structured or depend on the signal. 
Yet, autocorrelation analysis is easily adapted to different noise statistics. \TODO{Here, an example with colored noise would be of great help.}

@@

\TODO{need a lot of signal occurrences}

>>>>

The significance of the model bias in cryo--EM was stressed in~\cite{shatsky2009method,henderson2013avoiding} and demonstrated by the ``Einstein from noise'' example.
In this experiment, the image of Einstein is correlated with multiple images of pure noise. The noisy images are aligned to Einstein's image using cross-correlation and then averaged. 
In Figure~\ref{fig:EinsteinFromNoise} we conducted this experiment with different number of noise images. With merely 1000 noise images, Einstein's face is clearly observed in the averaged image. With $10,000$  image, the output is remarkably similar to the template image of Einstein rather than to pure noise. Crucially, the more noisy images we use, the more severe the model bias towards the template. 
In the context of cryo--EM, this example demonstrates how our prior assumptions about the particle may influence the reconstructed structure.
This true in general to any method which is based on template matching.

@@

	\caption{\label{fig:EinsteinFromNoise} Einstein from noise experiment. As a template, we used Einstein's image of size $50\times 50$ pixels. We drawn $n$ images of the same size whose entries are composed of i.i.d.\ Gaussian entries with the same mean and variance as Einstein's image. Each noise image was aligned with respect to the template by the maximal value of their cross correlation. After the alignment, all the images were averaged. As can be seen, the more pure noise images we use, the closer the averaged image to the template.   }	
\end{figure}

@@

Figure~\ref{fig:micro_example} shows a simple example of our problem with  one signal ($K=1$). 
Each panel demonstrates an image of size $250\times 250$, containing 4 occurrences of an Einstein image of size $50\times 50$ presented in Figure~\ref{fig:Einstein}. 
The left panel shows the data without noise. In the middle panel,  i.i.d.\ Gaussian noise with standard deviation $\sigma=0.5$  was added, however,  Einstein is still easily identified. In this scenario it is likely that standard detection algorithms can locate the copies of the signal within the micrograph. In the right panel, we show the same data swamped in  
i.i.d.\ Gaussian noise with standard deviation  $\sigma=3$.
Clearly, identifying the locations of Einstein in this micrograph is much more challenging. In fact, it can be shown that in the low signal--to--noise (\SNR) regime, 
detection  of individual image occurrences is impossible---even if the true image is known---and therefore particle picking is impossible.  This phenomenon is explained in detail in Appendix~\ref{sec:theory}.

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_clean}
		\caption{$\sigma = 0$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s05}
		\caption{$\sigma = 0.5$}
	\end{subfigure}
	\begin{subfigure}[h]{0.33\textwidth}
		\centering
		\includegraphics[scale=0.5]{micrograph_Einstein_example_s3}
		\caption{$\sigma = 3$}
	\end{subfigure}
	\caption{\label{fig:micro_example} Example of data sequences (micrographs) of size $250\times 250$ with additive i.i.d.\ Gaussian noise with variance $\sigma^2$. Each micrograph contains four occurrences of a $50 \times 50$ image of Einstein; see Figure~\ref{fig:Einstein}.}	
\end{figure}

@@

In this section we describe  some numerical experiments and show results. The experimental details and the algorithms are discussed at length in Section~\ref{sec:methods}.

In the first experiment, we estimated Einstein's image of size $50\times 50$ from multiple micrographs of size $4096\times 4096$ pixels, contaminated  with additive i.i.d.\ Gaussian noise with standard deviation $\sigma=3$ and $\SNR$ of $1/20$. An illustrative example of a micrograph appears in the right panel of Figure~\ref{fig:micro_example}.
Each micrograph contains multiple occurrences of Einstein's image  in random locations, while keeping a separation of  $49$ pixels between the images in each direction. From the micrographs, we estimated the signal's power spectrum, which is equivalent to the second-order autocorrelation. Then, in order to estimate the signal itself, we applied a standard phase retrieval algorithm called relaxed-reflect-reflect (RRR).

As an initial guess, we picked an image of the physicist Issac Newton; see Figure~\ref{fig:newton}. If the algorithm was prone to model bias, we would expect to get as an output an image that resembles Newton, similarly to the ``Einstein from noise'' effect of Figure~\ref{fig:EinsteinFromNoise}. However, the experiment exhibits the desired phenomenon: the more data we collect, the better the reconstruction quality.

Figure~\ref{fig:Einst_example} demonstrates several recovery results for different number of recorded micrographs. Figure~\ref{fig:error_per_micro} presents the normalized recovery error and the power spectrum estimation error as a function of the number of micrographs.
To measure the error, we use the root mean square error (RMSE) defined as 
\begin{equation}
\text{RMSE}  := \frac{\|x - \hat{x}\|_{\text{F}}}{\|x \|_{\text{F}}},
\end{equation} 
where $x$ and $\hat{x}$ are, respectively, the underlying and estimated image, and $\|\cdot\|_{\text{F}}$ stands for the Frobenius norm. 
As expected, the RMSE of estimating the power spectrum   decreases linearly with slope $-1/2$ in logarithmic scale. 
\TODO{Few comments: 1. We have a movie in the supplementary material. 2. Does the error also have the right slope? 3. We can put more images to exemplify the progress 4. Need to improve the micrograph's generation code  }

\begin{figure}[h!]
	\centering
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{Newton}
		\caption{\label{fig:newton}Newton  (model)}
	\end{subfigure}%
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{reconstruction2}
		\caption{\small $P =1024$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{reconstruction20}
		\caption{$P = 1024\times 10$}
	\end{subfigure}%
	\begin{subfigure}[h]{0.25\textwidth}
		\centering
		\includegraphics[scale=0.3]{reconstruction200}
		\caption{$P = 1024\times 100$}
	\end{subfigure}%
	\caption{\label{fig:Einst_example} Recovery of Einstein's image. The RRR algorithm is initialized with the image of the physicist Issac Newton in the left panel. We show estimations of Einstein using $P=1024,1024\times 10$ and $1024\times 100$ number of micrographs, each contains $700$ image occurrences on average and the $\SNR$ is $1/20$. The signal's estimation RMSE is $0.5514$, $0.3494$ and $0.2112$, respectively. The power spectrum estimation RMSE is, respectively, $0.0269 $, $0.0083$ and $0.0027$. \TODO{Can add more images to exemplify the progress}}	
\end{figure}

@@

Results from a similar experiment with three ($K=3$) one-dimensional signals are presented in Figure~\ref{fig:1Dheterosignals}.
 We define the ratio of the space occupied by the $i$th signal as 
\begin{equation}
\gamma_i = \frac{M_i L}{N},
\end{equation}
where $M_i$ is the number of signal's  occurrences, $L$ is the length of the signal and $N$ is the micrograph length.
These ratios present the {density} of the the signals in the data.
In the experiment, we do not assume to know these ratios, neither the noise level $\sigma$.

@@

\begin{figure}[h]
\centering
\includegraphics[scale=.7]{Einstein_progress}
\caption{\label{fig:error_per_micro} The left panel shows the RMSE of estimating the power spectrum of the signal from the data as a function of the number of  collected micrographs. As expected, the curve is linear with slope $-1/2$ in logarithmic scale. The right panel shows the RMSE  (in logarithmic scale) of recovering Einstein's image. \TODO{The program still runs so these figures will be extended to the right (more micrographs).}}	
\end{figure}

@@

\begin{figure}[t]
	\centering
	\includegraphics[width=\linewidth]{XP_1D_heterogeneous/hetero_example}
	\caption{An experiment for estimate three one-dimensional signals simultaneously with noise level $\sigma=3$. Red signals are the ground truth (targets) and the blue signals are our estimations. Individual RMSE of the estimates: $0.0239393 / 0.208925 / 0.0335956$. The estimated $\gamma$'s are $0.02574 / 0.01693 / 0.00818$ and the true ones $0.02561 / 0.01707 / 0.00853$. The $\SNR$ is $1/12$. \TODO{1. to replace with a ``progress'' plot 2. replace the triangle signal 3. add error progress figure}}
	\label{fig:1Dheterosignals}
\end{figure}

@@

Our results rely on two core assumptions that are not necessarily met by in an SPR experiment.  
First, we modeled the background information as i.i.d.\ additive noise. In practice, the background information may be structured or depend on the signal. 
Yet, autocorrelation analysis is easily adapted to different noise statistics.

@@
